|

### 2021-08-23 dissertation/msc_dissertation.Rmd
Further editing the dissertation

knitr::opts_chunk$set(echo = FALSE, fig.pos = 'H')

@@

 \newline

__ABSTRACT__ Numerous studies have show that analytical choices researchers make throughout the process of conducting a study affect the outcome. Recent crowdsourcing approaches to data analysis show that even well-intentioned analytical choices can make the difference between find a significant and strong or a non-significant effect. IN PROGRESS

@@

@@

 Crowdsourcing is one approach to better understand the influence of analysis strategies on research findings [@botvinik-nezer2020; @silberzahn2018]. A large number of researchers team up into smaller, independent groups to investigate the same research question based on the same dataset. For instance, in @silberzahn2018, 29 teams investigated the effects of a football player’s skin colour on the odds of being sent off the field. The variation between the analysis strategies was substantial. There were 29 different analyses with 21 different combinations of covariates. Twenty teams found a statistically significant effect of skin colour on the risk of being sent off the field. The authors also controlled for researchers’ prior believes and experience as well as peer-rated analysis quality, but these factors did not account for the variation of results. @botvinik-nezer2020 made similar observations. Such crowdsourcing approaches are well suited to show how different analysis strategies influence research findings, but they are extremely time and resource consuming. For instance, it took two and more than three years and 61 and 180 analysts, respectively, to perform the two studies mentioned above. Furthermore, although crowdsourcing approaches can cover more analysis strategies than a single conventional analysis, they are limited by the number of teams. They, therefore, cover a specific selection but not all possible strategies.

 Another approach to assess the effects of analysis strategies on research findings is multiverse analysis (also known as specification-curve analysis, [@simonsohn2020]. This approach has been proposed by @botvinik-nezer2020 and @silberzahn2018 as an alternative to crowdsourcing. In this approach, all reasonable analysis strategies are identified and performed. *Reasonable strategies* are defined as all statistically valid, non-redundant tests that are appropriate for the research question. The aggregated results of all reasonable analysis strategies are used to make inferences about the research question. Although the approach is statistically complex and computationally intense, it can be performed by one or a few researchers. Moreover, a researcher’s inherent strategy bias is neutralised, and noise is made transparent [@simonsohn2020]. Multiverse approaches are therefore well suited to better understand how different analysis strategies influence research results.

@@

# Analysis
## Analysis plan
     The project's objective was to define the space and the structure of all possible results of @silberzahn2018. The results space refers to the numerical interval between the lowest and highest possible outcome. It is created by running every possible analysis strategy. Hence, it is closely related to assessing the robustness of an effect. Robustness refers to the consistency of an effect under different model specifications. @patel2015 developed a standardised approach to assess an effect's robustness. I used this standardised approach to define the results space. The approach essentially comprises two parameters: the statistical models and the covariates (or control variables). In @silberzahn2018, the analysts used numerous statistical models like multiple linear regression, mixed-model logistic regression or Bayesian logistic regression. In total, there were 29 different modelling approaches. Additionally, each team used a different set of covariates. Across all teams, 15 covariates were used, resulting in $2^n$, i.e., $2^{15} = 32,768$ possible combinations of covariates. Adding all modelling possibilities to the equation yields a total of $29 * 2^{15} = 950,272$ possible analysis strategies. As this number of strategies exceeds the current project's resources, it was reduced to a more manageable number. Therefore, the project focused on the modelling approach that produced the median outcome of all analyses. The median, being the middle number of a given set of values, is a reasonable starting point to estimate the results space. However, even focusing on one analysis approach leaves $1 * 2^{15} = 32,768$ possibilities. Due to computational limitations, a random sample of 1,000 analysis strategies were conducted.

 In @silberzahn2018, the median outcome was produced by @team23, which will hereinafter be referred to as *team 23* due to its team number in the original study. *Team 23* first transformed the data and then conducted a mixed-model logistic regression. The first step of the present project was to replicate their transformation and analysis. Such replication can increase the confidence in the previous and the current approach. Moreover, it ensures that this project has the same starting point as *team 23*. As *team 23* made all scripts publicly available, this step was straightforward. The next step of the project was to define the results space and its structure. To this end, a random sample of covariates without replacement was drawn. “Without replacement” ensured that all covariate combinations were unique in the sample. The covariates were appended to the base (or core) variables. Base variables were those variables that were primarily assessed to answer the research question. In this case, the research question was whether a football player’s skin colour affects the odds of being sent off the field. *Team 23* defined two interaction terms as the base: “skin tone X implicit bias” and “skin tone X explicit bias.” (The variables are described in the “data” section.) Hence, all models had the following structure:
$$
\begin{aligned}
RedCard = SkinTone*ImplicitBias + SkinTone*ExplicitBias + CovariateCombination_{i}
\end{aligned}
$$ 
The relevant outcome parameters of each model were extracted. These parameters were the coefficient (i.e., effect) of skin tone, its standard error, test statistic and p-value. Similar to *Team 23* I calculated the 95% confidence intervals (CI). The estimates and their CIs were transformed to odds ratios (OR) through exponentiating them to the power of two. OR quantify the strength of association between two variables. An OR greater than one indicates that the dependent variable is more likely to occur given the independent variable, if it is lower than one it is less likely to occur. Eight OR outliers were excluded from visualisation. Four approached infinity, one was in the lower million range, one was 150, and a final one was 4.6.

@@

(ii) The raincloud plot @allen2021 combines a probability density plot, a box plot, and a scatter plot to give an unbiased, transparent view of the raw data. If the three plots are stacked vertically from top to bottom, they look like an eponymous cloud with rain drops. The strength of the raincloud plot is that it visualizes the raw data, summary statistics (median, 25th and 75th quartiles and CIs) and the probability density in a single plot. As the current project seeks to define and describe a results space, the raincloud plot is an ideal tool (for an example check Figure \ref{fig:rain_cloud_plot}). However, the plot does not allow to identify which covariates cause which outcomes. As the project also sought to understand the structure of the results space, it was essential to gain insights into how specific covariates influenced the outcomes. Therefore a third plot which complemented and extended the raincloud plot was developed.

@@

## Code
     This section provides a brief overview of the code and explains the reasoning behind it. First, the data transformation and analysis of *Team 23* [@team23] were replicated. The team made their project folder publicly available. It contains three scripts relevant to the current project: data exploration, transformation, and analysis. After duplicating their project folder on the local hard drive, all scripts ran without issues (only the working directories needed adjustment). The data exploration included the reasoning behind transforming the data and some cleaning. The data transformation restructured the data into a more intuitive format (more details on the topics of exploration and transformation are in the data section). The analysis script prepared the data by assigning variable classes (factors, numerical or boolean) and standardised a few variables, i.e. they were centred around the mean. Finally, the models were specified. The team ran both frequentist and Bayesian models. However, the current project focused on the frequentist approach.

 The code was based on @haessler2020 and @patel2015. The basic structure of the code is shown in the pseudocode table (see Algorithm \ref{alg:pseudocode}). Here, two important analytical decisions are briefly motivated. For more details, please refer to the commented code:

@@

\begin{algorithm}
\caption{Outlining high-level structure of the code. "Ran" in e.g. "FormulaRanEf" refers to "random-structure". "Ef" refers to "effect".}
\label{alg:pseudocode}
\SetKwInOut{Output}{Output}
\textbf{data} $\leftarrow$ prepared data based on team 23 \\
  \BlankLine
  $Variables$ $\leftarrow$ Define dependent ($dv$), base variables ($basevar$) and covariates ($covar$) \\
  $Specifications$ $\leftarrow$ Use $Variables$ to create matrix containing all possible covariate combinations \\
  \BlankLine
  $Formula$ $\leftarrow$ Paste $Specifications$ by row and append as column to $Specifications$ \\
  $FormulaRanEf; FormulaEf$ $\leftarrow$ Separate formulas based on including a random-structure \\
  \BlankLine
  $SampleRanEf; SampleEf$ $\leftarrow$ Unique, random sample from $FormulaRanEf; FormulaEf$ \\
  \BlankLine
  $ExtractorFunction$ $\leftarrow$ Extracting model statistics and store them in object \\
  \BlankLine
  $OutputsRanEf$ $\leftarrow$ \textbf{lapply(data, $ExtractorFunction$(glmer($dv$, x = $basevar + SampleRanEf$)))} \\
  $OutputsRanEfEF$ $\leftarrow$ \textbf{lapply(data, $ExtractorFunction$(glm($dv$, x = $basevar + SampleEf$)))} \\
  \BlankLine
  $ModelOutputs$ $\leftarrow$ Merge $OutputsRanEf$ and $OutputsEF$ \\
  $ModelOutputs$ $\leftarrow$ Compute odds ratios and confidence intervals \\
  \BlankLine
  $JoinedData$ $\leftarrow$ \textbf{LeftJoin($Specifications$, $ModelOutputs$, by = $Formula$)} \\
  \BlankLine
  \Output{Rain cloud plot}
  \BlankLine
  $CovariateEffects$ $\leftarrow$ \textbf{LinearModel($OddsRatios$, x = $Specifications$, data = $JoinedData$)} \\
  \BlankLine
  \Output{Covariate effects plot}
  \Output{SCA plot, sorted by $CovariateEffects$}
  \BlankLine
  \Output{Volcano plot}
\end{algorithm}

 The code is written in R (Version 1.4.1103) on macOS Big Sur (Version 11.4) and can be retrieved from my [GitHub repository](https://github.com/sebastianplnr/msc_dissertation_project). The code from *Team 23* [@team23] can be retrieved from their [OSF repository](https://osf.io/akqt4/). Following R packages were used *here 1.0.1* [@here], *data.table 1.14.0* [@datatable], *tidyverse 1.3.1* [@tidyverse], *lme4 1.1-27.1* [@lme4], *pbmcapply 1.5.0* [@pbmcapply], *PupillometryR 0.0.3* [@PupillometryR] and *cowplot 1.1.1* [@cowplot].

## Data description and preparation
     The data was retrieved from @silberzahn2018. It contained information about football players, their encounters with referees and the received cards (yellow, yellow/red and red). Moreover, it included a player’s position, age, club, league country, victories, ties, defeats, and goals. In addition, a skin tone rating based on two independent judges was included. The referees were numerically coded to protect their identity. The referees’ countries of origin were also included as well as implicit and explicit racism bias scores for their respective countries. The exploratory data analysis (EDA) of *Team 23* showed that each row of the dataset represents a unique player-referee combination listing all their encounters as well as the couple’s total number of received/assigned cards. *Team 23* stated that it preferred a different data format where each player-referee encounter is reflected by one row. This way, each encounter had a maximum of one red card. To achieve this format, the data had to be transformed, i.e. disaggregated. Further EDA showed that receiving a red card was highly unlikely ($p = 0.008%$), i.e. the data were highly skewed. The team, therefore, used a logistic regression which is a statistical modelling technique equipped to deal with skewed/binary distributions. Figure \ref{fig:var_plots} shows an overview of the properties of the core variables used in their analysis.

 Finally, the team excluded all referees who did not have at least 22 encounters with players. Every football game includes (at least) 22 players, 11 players per team. If a referee has less than 22 player-encounters, there are missing cases. According to team 23, this was mostly the case in referees officiating games of minor leagues. As they focused on the major European leagues (England, Germany, Italy, and Spain), referees with less than 22 player-encounters were excluded. This step excluded about 66% of the referees but retained 97.4% of all player-referee combinations. The final dataset used for the current project contained 335,537 observations and 19 variables.  \newline

```{r var_plots, echo = FALSE, fig.width = 8, fig.align = "center", fig.cap = "Team 23 base variables properties. A) There was a strong skew towards receiving no red card compared to receiving a red card. B) Most players had a lower skin tone rating, i.e., a brighter skin colour. C) The implicit racial bias scores hardly varied. D) The explicit racial bias scores varied slightly more than the implicit scores. (The latter two were centred around the mean; hence, zero does not mean there is no bias.)"}

@@

# Results
## Replicating *Team 23* analysis
     First, the analysis of *Team 23* was replicated using their scripts. Skin tone significantly affected the odds of being sent off the field ($OR = 1.311$, 95%CI [1.099, 1.563], $p = 0.003$). This means that when keeping all other variables constant, for every unit increase in skin tone rating (darker skin tone), the odds of being sent off the field increased by about 131%. The interaction terms of skin tone and implicit racial bias ($OR = 0.004$, 95%CI [0.000, 23.259], $p = 0.211$) as well as skin tone and explicit racial bias ($OR = 1.837$, 95%CI [0.493, 6.848], $p = 0.365$) were both non-significant. In accordance with the original analysis concerning additional variables, there were significant differences of OR between leagues and positions as well as implicit racial bias scores. Explicit bias scores were on the verge of being non-significant. Thus, the results of team 23 were reproducible without any adjustment.

## Exploring the results space - specification curve
     Second, the specification curve was calculated to visualise the results space (Figure \ref{fig:sca_plot}). The horizontal axes of the top and bottom chart show the specifications sorted by their OR from lowest to highest. The OR are shown on the vertical axis of the top plot. The points in the top plot (which together look like a line) each represent the outcome of one statistical model. Black and red points refer to statistically significant and non-significant outcomes, respectively. Overall, 89.7% of the outcomes were significant. Except for the lowest outcome ($OR_{lowest} = 1.080$) all points were close to their neighbours. This indicates that none of the specified models was responsible for a sudden increase in the outcome measure. The bottom table lists all covariates on its vertical axis, sorted from most positive to most negative impact. For instance, the covariate player had the most substantial positive impact, while the covariate club had the most substantial negative impact. Each column of the table represents one mode. A coloured cell indicates the presence of the variable, an uncoloured cell its absence. Again, black and red colours indicate that the outcome measure was significant or non-significant, respectively.

 The top and bottom plots work in tandem. Each outcome (i.e., point) in the top corresponds to the indicated covariates in the bottom table. Regular specification curve analyses focus on all *reasonable* specifications; hence, the number of specifications is much lower than in the present study. With these lower numbers of specifications, it is possible to tease the different model specifications apart. However, the current project’s goal was to define the whole results space, i.e. all *possible* specifications. With such a high number of specifications, it is no longer possible to tease the different specifications apart. It is, however, interesting to look at the top and bottom rows of the table. The top row shows that the covariate player was particularly often included in models with a higher outcome. Conversely, the covariate club was particularly often included in models with a lower outcome. \newline

```{r sca_plot, echo = FALSE, fig.width = 8, fig.align = "center", fig.cap = "Specification curve. Results space ranges from about 1.05 to 1.4. However, due to the large number of specifications, the specification curve does not allow for identifying the relevance of the different covariates.", warning = FALSE}

@@

## Exploring the results space – raincloud plot
     Third, the results space was further explored by calculating raincloud plots (Figure \ref{fig:rain_cloud_plot}). The horizontal axis shows the OR. From top to bottom, the three components show the probability density distribution, the box plot and the raw data. Based on the raw data points, the results space can be defined as the interval between 1.081 and 1.383. The 1st quartile was 1.206, the median 1.248 and the 3rd quartile 1.293. Thus, the interquartile range where 50% of the data were included was relatively small (1.206-1.293), indicating a low statistical dispersion. The dashed line indicates the original results of *Team 23*, which has been the median outcome of all analysis strategies in @silberzahn2018. The outcome of *Team 23* (1.31) was included in the results space but outside the middle 50% of the data. The CI calculated by team 23 ranged from 1.10 to 1.56. The lower bound of their CI was close to the lower margin of the results space of the present study. In contrast, the upper bound of their CI was notably higher than the present results space. This suggests a potential overestimating of the variability of the results in the original analysis.

 The probability density distribution reveals another exciting feature of the data. The distribution has two peaks which suggest two latent covariate structures whose outcomes centre around two points. Those two focal points were located at about 1.13 and 1.19. Based on this graph alone, it is impossible to decipher the latent structures, i.e. determine the responsible covariates or combinations of covariates. To test the hypothesis of two latent structures, it would be interesting to observe how the distribution evolves when more specifications are run. In the presence of latent structures, the peaks should get more pronounced. In the absence of latent structures, the peaks should smoothen out, and the distribution should approach a normal distribution. \newline

```{r rain_clod_plot, echo = FALSE, fig.width = 8, fig.cap = "Raincloud plot. The results space ranges from about 1.05 to 1.4. The distribution showed two peaks which suggests an influential latent covariate structure. The original median effect was within the results space but outside the middle 50% of the data."}

@@

## Specifying covariate effects – scatter plot 
     Fourth, to better understand the relevance of the covariates, another scatter plot was calculated (Figure \ref{fig:covar_plot}). This graph shows the specified covariate effects. On the horizontal axis are the covariates; on the vertical axis are their estimated effects. The covariates are sorted from negative to positive impact. Black indicates a significant effect, red non-significant. The error bars represent the 95% CI. Four out of 15 covariates were significant. All effect sizes, including the significant covariates’, were relatively weak. The CIs were relatively small, which was likely due to the large sample size. The OR of the model without any covariates (i.e., core variables only) was 1.262. Adding the sum of all positive covariate effects (games + goals + referee country + player, respectively) to the base model resulted in a maximum possible effect of 1.357. The maximum observed OR was 1.382, however. This difference suggests that there were more impactful covariate combinations or interactions that have not yet been identified. Vice versa, subtracting the sum of all negative covariate effects from the base model resulted in an OR of 1.070, whereas the minimum observed effect was an OR of 1.081. This suggests that there were also covariate combinations or interactions with a higher negative impact. \newline

```{r covar_plot, echo = FALSE, fig.width = 8, fig.align = "center", fig.cap = "Scatter plot. Four out of 15 covariates were significant, although their effects were relatively weak. This suggests that covariates interactions were responsible for the higher outcome values."}

@@

## Exploring the results structure – volcano plot
     Lastly, another graph was produced despite not being planned. Therefore, it is only used to strengthen previous points and probe future research briefly. The results above suggest one or more meaningful latent covariate interactions; these were further explored. Although @patel2015 had a different objective, their methodology was similar to the current project. They visualised the outcomes as a so-called volcano plot (see Figure \ref{fig:rain_cloud_plot}), showing the relationship between the p-values and effect sizes. A similar graph was calculated for the current project. The two distinct “streams” provided further support for two distinct underlying constructs. Future research might specify which covariate combination underlies this structure.

>>>>

knitr::opts_chunk$set(echo = FALSE, fig.width = 8, fig.align = "center", fig.pos = 'H')

@@

@@

# Abstract
Crowdsourcing and multiverse analysis usually only assess reasonable analysis strategies. Little is known about how the results space of these *reasonable* analyses relates to the results space of all *possible* analyses. The current project, therefore, aimed at defining and understanding the results space and structure, respectively, of all possible analyses for @silberzahn2018. Due to time and resources limitations, the project focused on the analysis producing the median outcome and randomly sampled 1,000 covariate combinations out of all possible ones. The findings indicate that the full results space is narrower than the original confidence intervals suggest. If this were to be true for all analyses conducted in @silberzahn2018, the variability between the analysis strategies were likely smaller than previously though. Future studies have to confirm this hypothesis by first running the "median model" with all covariate combinations and for each statistical model.

@@

 Crowdsourcing is one approach to better understand the influence of analysis strategies on research findings [@botvinik-nezer2020; @silberzahn2018]. A large number of researchers team up into smaller, independent groups to investigate the same research question based on the same dataset. For instance, in @silberzahn2018, 29 teams investigated the effects of a football player’s skin colour on the odds of being sent off the field. The variation between the analysis strategies was substantial. There were 29 different analyses with 21 different combinations of covariates. Twenty teams found a statistically significant effect of skin colour on the odds of being sent off the field. The authors also controlled for researchers’ prior believes and experience as well as peer-rated analysis quality, but these factors did not account for the variation of results. @botvinik-nezer2020 made similar observations. Such crowdsourcing approaches are well suited to show how different analysis strategies influence research findings, but they are extremely time and resource consuming. For instance, it took two and more than three years and 61 and 180 analysts, respectively, to perform the two studies mentioned above. Furthermore, although crowdsourcing approaches can cover more analysis strategies than a single conventional analysis, they are limited by the number of teams. They, therefore, cover a specific selection but not all possible strategies.

 Another approach to assess the effects of analysis strategies on research findings is multiverse analysis [also known as specification-curve analysis, @simonsohn2020]. This approach has been proposed by @botvinik-nezer2020 and @silberzahn2018 as an alternative to crowdsourcing. In this approach, all reasonable analysis strategies are identified and performed. *Reasonable strategies* are defined as all statistically valid, non-redundant tests that are appropriate for the research question. The aggregated results of all reasonable analysis strategies are used to make inferences about the research question. Although the approach is statistically complex and computationally intense, it can be performed by one or a few researchers. Moreover, a researcher’s inherent strategy bias is neutralised, and noise is made transparent [@simonsohn2020]. Multiverse approaches are therefore well suited to better understand how different analysis strategies influence research results.

@@

# Analysis
## Analysis plan
     The project's objective was to define the space and the structure of all possible results of @silberzahn2018. The results space refers to the numerical interval between the lowest and highest possible outcome. It is created by running every possible analysis strategy. Hence, it is closely related to assessing the robustness of an effect. Robustness refers to the consistency of an effect under different model specifications. @patel2015 developed a standardised approach to assess an effect's robustness. This standardised approach to define the results space was used. The approach essentially comprises two parameters: the statistical models and the covariates (or control variables). In @silberzahn2018, the analysts used numerous statistical models like multiple linear regression, mixed-model logistic regression or Bayesian logistic regression. In total, there were 29 different modelling approaches. Additionally, each team used a different set of covariates. Across all teams, 15 covariates were used, resulting in $2^n$, i.e., $2^{15} = 32,768$ possible combinations of covariates. Adding all modelling possibilities to the equation yields a total of $29 * 2^{15} = 950,272$ possible analysis strategies. As this number of strategies exceeds the current project's resources, it was reduced to a more manageable number. Therefore, the project focused on the modelling approach producing the median outcome of all analyses. The median, being the middle number of a given set of values, is a reasonable starting point to estimate the results space. However, even focusing on one analysis approach leaves $1 * 2^{15} = 32,768$ possibilities. Due to computational limitations, a random sample of 1,000 analysis strategies were conducted.

 In @silberzahn2018, the median outcome was produced by @team23, which will hereinafter be referred to as *Team 23* due to its team number in the original study. *Team 23* first transformed the data and then conducted a mixed-model logistic regression. The first step of the current project was to replicate their transformation and analysis. Such replication can increase the confidence in the previous and the current approach. Moreover, it ensures that this project has the same starting point as *Team 23*. As team made all scripts publicly available, this step was straightforward. The next step of the project was to define the results space and its structure. To this end, a random sample of covariates without replacement was drawn. “Without replacement” ensured that all covariate combinations were unique in the sample. The covariates were appended to the base (or core) variables. Base variables were those variables that were primarily assessed to answer the research question. In this case, the research question was whether a football player’s skin colour affects the odds of being sent off the field. *Team 23* defined two interaction terms as the base: “skin tone X implicit bias” and “skin tone X explicit bias.” (The variables are described in the “data” section.) Hence, all models had the following structure:
$$
\begin{aligned}
RedCard = SkinTone*ImplicitBias + SkinTone*ExplicitBias + CovariateCombination_{i}
\end{aligned}
$$ 
The relevant outcome parameters of each model were extracted. These parameters were the coefficient (i.e., effect) of skin tone, its standard error, test statistic and p-value. Similar to *Team 23* the 95% confidence intervals (CI) were calculated. The estimates and their CIs were transformed to odds ratios (OR) through exponentiating them to the power of two. OR quantify the strength of association between two variables. An OR greater than one indicates that the dependent variable is more likely to occur given the independent variable, if it is lower than one it is less likely to occur. Eight OR outliers were excluded from visualisation. Four approached infinity, one was in the lower million range, one was 150, and a final one was 4.6.

@@

(ii) The raincloud plot [@allen2021] combines a probability density plot, a box plot, and a scatter plot to give an unbiased, transparent view of the raw data. If the three plots are stacked vertically from top to bottom, they look like an eponymous cloud with rain drops. The strength of the raincloud plot is that it visualizes the raw data, summary statistics (median, 25th and 75th quartiles and CIs) and the probability density in a single plot. As the current project seeks to define and describe a results space, the raincloud plot is an ideal tool (for an example check Figure \ref{fig:rain_cloud_plot}). However, the plot does not allow to identify which covariates cause which outcomes. As the project also sought to understand the structure of the results space, it was essential to gain insights into how specific covariates influenced the outcomes. Therefore a third plot which complemented and extended the raincloud plot was developed.

@@

## Code
     This section provides a brief overview of the code and explains the reasoning behind it. First, the data transformation and analysis of *Team 23* [@team23] were replicated. The team made their project folder publicly available. It contains three scripts relevant to the current project: data exploration, transformation, and analysis. After duplicating their project folder on the local hard drive, all scripts ran without issues (only the working directories needed adjustment). The data exploration included the reasoning behind transforming the data and some cleaning. The data transformation restructured the data into a more intuitive format (more details on the topics of exploration and transformation are in the "data" section). The analysis script prepared the data by assigning variable classes (factors, numerical or boolean) and standardised a few variables, i.e. they were centred around the mean. Finally, the models were specified. The team ran both frequentist and Bayesian models. However, the current project focused on the frequentist approach.

 The code was based on @haessler2020 and @patel2015. The basic structure of the code is shown in the pseudocode table (see Pseudocode \ref{alg:pseudocode}). Here, two important analytical decisions are briefly motivated. For more details, please refer to the commented code:

@@

\begin{algorithm}
\SetAlgorithmName{Pseudocode}{pseudocode}{List of pseudocodes}
\caption{Outlining high-level structure of the code. "Ran" in e.g. "FormulaRanEf" refers to "random-structure". "Ef" refers to "effect".}
\label{alg:pseudocode}
\SetKwInOut{Output}{Output}
\textbf{data} $\leftarrow$ prepared data based on team 23 \\
  \BlankLine
  $Variables$ $\leftarrow$ Define dependent ($dv$), base variables ($basevar$) and covariates ($covar$) \\
  $Specifications$ $\leftarrow$ Use $Variables$ to create matrix containing all possible covariate combinations \\
  \BlankLine
  $Formula$ $\leftarrow$ Paste $Specifications$ by row and append as column to $Specifications$ \\
  $FormulaRanEf; FormulaEf$ $\leftarrow$ Separate formulas based on including a random-structure \\
  \BlankLine
  $SampleRanEf; SampleEf$ $\leftarrow$ Unique, random sample from $FormulaRanEf; FormulaEf$ \\
  \BlankLine
  $ExtractorFunction$ $\leftarrow$ Extracting model statistics and store them in object \\
  \BlankLine
  $OutputsRanEf$ $\leftarrow$ \textbf{lapply(data, $ExtractorFunction$(glmer($dv$, x = $basevar + SampleRanEf$)))} \\
  $OutputsEf$ $\leftarrow$ \textbf{lapply(data, $ExtractorFunction$(glm($dv$, x = $basevar + SampleEf$)))} \\
  \BlankLine
  $ModelOutputs$ $\leftarrow$ Merge $OutputsRanEf$ and $OutputsEf$ \\
  $ModelOutputs$ $\leftarrow$ Compute odds ratios and confidence intervals \\
  \BlankLine
  $JoinedData$ $\leftarrow$ \textbf{LeftJoin($Specifications$, $ModelOutputs$, by = $Formula$)} \\
  \BlankLine
  \Output{Rain cloud plot}
  \BlankLine
  $CovariateEffects$ $\leftarrow$ \textbf{LinearModel($OddsRatios$, x = $Specifications$, data = $JoinedData$)} \\
  \BlankLine
  \Output{Covariate effects plot}
  \Output{SCA plot, sorted by $CovariateEffects$}
  \BlankLine
  \Output{Volcano plot}
\end{algorithm}

 The code is written in R (Version 1.4.1103) on macOS Big Sur (Version 11.4) and can be retrieved from my [GitHub repository](https://github.com/sebastianplnr/msc_dissertation_project). The code from *Team 23* [@team23] can be retrieved from their [OSF repository](https://osf.io/akqt4/). Following R packages were used: *here 1.0.1* [@here], *data.table 1.14.0* [@datatable], *tidyverse 1.3.1* [@tidyverse], *lme4 1.1-27.1* [@lme4], *pbmcapply 1.5.0* [@pbmcapply], *PupillometryR 0.0.3* [@PupillometryR] and *cowplot 1.1.1* [@cowplot].

## Data description and preparation
     The data was retrieved from @silberzahn2018. It contained information about football players, their encounters with referees and the received cards (yellow, yellow/red and red). Moreover, it included a player’s position, age, club, league country, victories, ties, defeats, and goals. In addition, a skin tone rating based on two independent judges was included. The referees were numerically coded to protect their identity. The referees’ countries of origin were also included as well as implicit and explicit racism bias scores for their respective countries. The exploratory data analysis (EDA) of *Team 23* showed that each row of the dataset represents a unique player-referee combination listing all their encounters as well as the couple’s total number of received/assigned cards. *Team 23* stated that it preferred a different data format where each player-referee encounter is reflected by one row. This way, each encounter had a maximum of one red card. To achieve this format, the data had to be transformed, i.e. disaggregated. Further EDA showed that receiving a red card was highly unlikely (0.8%), i.e. the data were highly skewed. The team, therefore, used a logistic regression which is a statistical modelling technique equipped to deal with skewed/binary distributions. Figure \ref{fig:var_plots} shows an overview of the properties of the core variables used in their analysis.

 Finally, the team excluded all referees who did not have at least 22 encounters with players. Every football game includes (at least) 22 players, 11 players per team. If a referee has less than 22 player-encounters, there are missing cases. According to *Team 23*, this was mostly the case in referees officiating games of minor leagues. As they focused on the major European leagues (England, Germany, France, and Spain), referees with less than 22 player-encounters were excluded. This step excluded about 66% of the referees but retained 97.4% of all player-referee combinations. The final dataset used for the current project contained 335,537 observations and 19 variables.  \newline

```{r var_plots, fig.cap = "Team 23 base variables properties. A) There was a strong skew towards receiving no red card compared to receiving a red card. B) Most players had a lower skin tone rating, i.e., a brighter skin colour. C) The implicit racial bias scores hardly varied. D) The explicit racial bias scores varied slightly more than the implicit scores. (The latter two were centred around the mean; hence, zero does not mean there is no bias.)"}

@@

# Results
## Replicating *Team 23* analysis
     First, the analysis of *Team 23* was replicated using their scripts. Skin tone significantly affected the odds of being sent off the field ($OR = 1.311$, 95%CI [1.099, 1.563], $p = 0.003$). This means that when keeping all other variables constant, for every unit increase in skin tone rating (darker skin tone), the odds of being sent off the field increased by about 131%. The interaction terms of skin tone and implicit racial bias ($OR = 0.004$, 95%CI [0.000, 23.259], $p = 0.211$) as well as skin tone and explicit racial bias ($OR = 1.837$, 95%CI [0.493, 6.848], $p = 0.365$) were both non-significant. In accordance with the original analysis concerning additional variables, there were significant differences of OR between leagues and positions as well as implicit racial bias scores. Explicit bias scores were on the verge of being non-significant. Thus, the results of *Team 23* were reproducible without any adjustment.

## Exploring the results space - specification curve
     Second, the specification curve was calculated to visualise the results space (Figure \ref{fig:sca_plot}). The horizontal axes of the top and bottom chart show the specifications sorted by their OR from lowest to highest. The OR are shown on the vertical axis of the top plot. The points in the top plot (which together look like a line) each represent the outcome of one statistical model. Black and red points refer to statistically significant and non-significant outcomes, respectively. Overall, 89.7% of the outcomes were significant. Except for the lowest outcome ($OR_{lowest} = 1.081$) all points were close to their neighbours. This indicates that none of the specified models was responsible for a sudden increase in the outcome measure. The bottom table lists all covariates on its vertical axis, sorted from most positive (top) to most negative impact (bottom). For instance, the covariate "player" had the most substantial positive impact, while the covariate "club" had the most substantial negative impact. Each column of the table represents one model. A coloured cell indicates the presence of the variable, an uncoloured cell its absence. Again, black and red colours indicate that the outcome measure was significant or non-significant, respectively.

 The top and bottom plots work in tandem. Each outcome (i.e., point) in the top corresponds to the indicated covariates in the bottom table. Regular specification curve analyses focus on all *reasonable* specifications; hence, the number of specifications is much lower than in the present study. With these lower numbers of specifications, it is possible to tease the different model specifications apart. However, the current project’s goal was to define the whole results space, i.e. all *possible* specifications. With such a high number of specifications, it is no longer possible to tease the different specifications apart. It is, however, interesting to look at the top and bottom rows of the table. The top row shows that the covariate "player" was particularly often included in models with a higher outcome. Conversely, the covariate "club" was particularly often included in models with a lower outcome. \newline

```{r sca_plot, fig.cap = "Specification curve. Results space ranges from about 1.1 to 1.4. However, due to the large number of specifications, the specification curve does not allow for identifying the relevance of the different covariates.", warning = FALSE}

@@

## Exploring the results space – raincloud plot
     Third, the results space was further explored by calculating raincloud plots (Figure \ref{fig:rain_cloud_plot}). The horizontal axis shows the OR. From top to bottom, the three components show the probability density distribution, the box plot and the raw data. Based on the raw data points, the results space can be defined as the interval between 1.081 and 1.383. The 1st quartile was 1.206, the median 1.248 and the 3rd quartile 1.293. Thus, the interquartile range where 50% of the data were included was relatively narrow (1.206 to 1.293), indicating a low statistical dispersion. The dashed line indicates the original result of *Team 23*, which has been the median outcome of all analysis strategies in @silberzahn2018. The outcome of *Team 23* (1.31) was included in the results space but outside the middle 50% of the data. The CI calculated by team 23 ranged from 1.10 to 1.56. The lower bound of their CI was close to the lower margin of the results space of the present study. In contrast, the upper bound of their CI was notably higher than the present results space. This suggests a potential overestimating of the variability of the results in the original analysis.

 The probability density distribution reveals another interesting feature of the data. The distribution has two peaks which suggest two latent covariate structures whose outcomes centre around two points. Those two focal points were located at about 1.13 and 1.19. Based on this graph alone, it is impossible to decipher the latent structures, i.e. determine the responsible covariates or combinations of covariates. To test the hypothesis of two latent structures, it would be interesting to observe how the distribution evolves when more specifications are run. In the presence of latent structures, the peaks should get more pronounced. In the absence of latent structures, the peaks should smoothen out, and the distribution should approach a normal distribution. \newline

```{r rain_cloud_plot, fig.cap = "Raincloud plot. The results space ranges from about 1.08 to 1.38. The distribution showed two peaks which suggests an influential latent covariate structure. The original median effect was within the results space but outside the middle 50 percent of the data."}

@@

## Specifying covariate effects – scatter plot 
     Fourth, to better understand the relevance of the covariates, another scatter plot was calculated (Figure \ref{fig:covar_plot}). This graph shows the specified covariate effects. On the horizontal axis are the covariates; on the vertical axis are their estimated effects. The covariates are sorted from negative to positive impact. Black indicates a significant effect, red non-significant. The error bars represent the 95% CI. Four out of 15 covariates were significant. All effect sizes, including the significant covariates’, were relatively weak. The CIs were relatively small, which was likely due to the large sample size. The OR of the model without any covariates (i.e., core variables only) was 1.262. Adding the sum of all positive covariate effects (games + goals + referee country + player, respectively) to the base model resulted in a maximum possible effect of 1.357. The maximum observed OR was 1.382, however. This difference suggests that there were more impactful covariate combinations or interactions that have not yet been identified. Vice versa, subtracting the sum of all negative covariate effects from the base model resulted in an OR of 1.070, whereas the minimum observed effect was an OR of 1.081. This suggests that there were also covariate combinations with a higher negative impact. \newline

```{r covar_plot, fig.cap = "Scatter plot. Four out of 15 covariates were significant, although their effects were relatively weak. This suggests that covariates interactions were responsible for the higher outcome values."}

@@

## Exploring the results structure – volcano plot
     Lastly, another graph was produced despite not being planned. Therefore, it is only used to strengthen previous points and probe future research. The results above suggest one or more meaningful latent covariate combinations; these were further explored. Although @patel2015 had a different objective, their methodology was similar to the current project. They visualised the outcomes as a so-called volcano plot (see Figure \ref{fig:rain_cloud_plot}), showing the relationship between the p-values and effect sizes. A similar graph was calculated for the current project. The two (potentially three) distinct “streams” provided further support for two (or three) distinct underlying constructs. Future research might specify which covariate combinations underly these structures.
